---
title: "Analysis of SensorLog and GT3X Data"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Analysis of SensorLog and GT3X Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(waterways)
library(lubridate)
library(tidyr)
```


# Introduction
In this vignette, we will go through the steps for analyzing SensorLog data. We will cover reading in the data, checking for potential duplicate timestamps, calculating distance metrics from a fixed point (commonly a participant's home), and creating daily summaries.

We will also integrate ActiGraph data to show activity at different distances.

# Methods
## Data
The data used is example data embedded in the waterways package. It is a single participant for approximately two hours' worth of GPS and activity data. The SensorLog data is stored in a zip file and contains a single CSV file. The data is stored in the `extdata` folder of the package. We can get the path of the file on each user's machine using `system.file` and the package name:

```{r sensorlog_file}
library(waterways)
file = system.file(
  "extdata", "SensorLogFiles_my_iOS_device_250311_14-55-58.zip",
  package = "waterways")
file
```

This can also be done using the wrapper from `waterways`:
```{r ww_sensorlog_file}
ww_example_sensorlog_file()
```

### Reading in the data
The data is read in using the `ww_read_sensorlog` function:

```{r read_sensorlog}
library(dplyr)
df = ww_read_sensorlog(file, robust = FALSE)
df = df %>% select(-file) # we don't need to see which file this came from
head(df)
```

The `robust` option will run `rewrite_sensorlog_csv` on the data. The `rewrite_sensorlog_csv` function attempts to fix shifts in the data, which have been seen due to a bug in the iOS app. The default for this option is `FALSE` because reading and writing to a temporary file can take much longer. This function will also flag any values of latitude and longitude that are 0 (absolute value $< 0.00001$). This flag is created because the iOS app will sometimes record 0 for latitude and longitude when the GPS signal is lost.

We see that the data has the following columns, which are renamed from the original data.
```{r}
colnames(df)
```
The mapping from the original column names to the new column names can be seen using `ww_sensorlog_csv_colnames_mapping`:
```{r}
ww_sensorlog_csv_colnames_mapping()
```


### Processing the data
The data is then processed using the `ww_process_sensorlog` function. This function (currently) will:

1. Check the data for duplicated data using `ww_check_data`.
2. Calculate distance from a central point if provided using `ww_calculate_distance`
3. Process the timing data using `ww_process_time`, including checking for duplicate times. This function will also determine whether the data is in the correct timezone by estimating the expected timezone (in the `expected_timezone` field, set to `America/New_York`) based on the GPS coordinates, using the `lutz::tz_lookup_coords` function. If this check is not desired, set `expected_timezone = NULL`:

```{r process_sensorlog_fail, error=TRUE}
ww_process_sensorlog(df)
```

As the error indicates, there are some duplicate times in this file. We can look at those specific duplicate times and see that even to 3 digits for milliseconds, there are multiple measurements:

```{r}
df %>% 
  add_count(time) %>% 
  filter(n > 1) %>% 
  select(time, lat, lon, starts_with("accel"))
```

Thus, we do not want to enforce unique times, so we will set `check_data = FALSE`:

```{r process_sensorlog}
df_proc = ww_process_sensorlog(df, check_data = FALSE, apply_tz = TRUE)
df_proc
```

Note that the `apply_tz` argument will apply the time zone of the data. This is helpful because it sets the correct time zone of the data in absolute terms. Unfortunately, this means researchers have to manually shift time-based filters such as hours. For example, if you want to find data between 10 AM and 6 PM in the EST time zone, you need to shift those values to GMT/UTC. This also can shift records into different dates. The data is likely sleep or rest data, but this can skew summary numbers. For analysis, we will not apply the timezone:

```{r process_sensorlog_for_real}
df = ww_process_sensorlog(df, check_data = FALSE, apply_tz = FALSE)
df
```



The data times will be projected into the `GMT` timezone to agree with the output from ActiGraph and `read.gt3x::read.gt3x`. **NOTE:** this shifts times from the original timezone. For example, if the data was at "2025-03-25 14:14:00" with a timezone of `-04:00`, then the data will be in the `GMT` timezone, but the printed time will be "2025-03-25 18:14:00 GMT". You can set the timezone to the correct timezone using `lubridate::with_tz` or `lubridate::force_tz` (they have different behavior).

You can also pass in the `tz = "EST"` argument to `ww_process_sensorlog`, which is passed to `ww_process_time`. An additional `char_time` column is added so that the original character values of the time are retained. This lets users check timezone conversions, which can lead to issues with merging data, determining windows (e.g. wake), or other time-based analyses.

```{r}
df %>% 
  select(time, char_time, timestamp) %>% 
  head()
```



### Calculating distance
We did not pass in the latitude and longitude of a fixed point when calling the processing function, but we could have. We can also calculate the distance separately using `ww_calculate_distance`. First, we need the latitude and longitude of the fixed point. We can use the `tidygeocoder` package to get the coordinates for the Johns Hopkins Bloomberg School of Public Health, where this data was collected. We will use the address `615 N Wolfe St, Baltimore MD`, using the [Census Geocoder](https://geocoding.geo.census.gov/geocoder/) for free:

```{r geo_run, echo = FALSE}
geo = structure(list(
  address = "615 N Wolfe St, Baltimore MD", lat = 39.297622492942, 
  long = -76.590753361728), class = c("tbl_df", "tbl", "data.frame"
  ), row.names = c(NA, -1L))
print(geo)
lat = geo$lat
lon = geo$long
```

The code below shows how to geocode the address live, but the vignette uses
the precomputed object above so it builds without network access.

```{r geo_show, eval = FALSE}
geo = tidygeocoder::geo("615 N Wolfe St, Baltimore MD", method = "census")
print(geo)
lat = geo$lat
lon = geo$long
```

The next section shows how to geocode and query the EPA walkability index
interactively. The vignette does not run these network calls so that it can
build in isolated environments.

```{r census_geocode, eval = FALSE}
df_census = data.frame(street = "615 N Wolfe St",
                       city = "Baltimore", state = "MD",
                       zip = "21205") %>%
  censusxy::cxy_geocode(
    street = "street",
    city = "city",
    state = "state",
    zip = "zip",
    output = "full",
    return = "geographies",
    benchmark = "Public_AR_Current",
    vintage = "Census2010_Current")
df_census = df_census %>%
  mutate(GEOID10 = ww_fips12(cxy_state_id, cxy_county_id, cxy_tract_id, cxy_block_id))
epa_walkability = ww_epa_walkability(geoid = df_census$GEOID10)
print(epa_walkability)
```

```{r get_epa_cat, eval = FALSE}
print(epa_walkability %>%
        as.data.frame() %>%
        select(NatWalkInd, cat_walk_index))
```



We can then calculate the distance from this fixed point using the `ww_calculate_distance` function:

```{r calc_distance}
df = ww_calculate_distance(df,
                           lat = lat,
                           lon = lon)
```

This function uses `geosphere::distm` to calculate the distance in meters from the fixed point. We can see the new columns that were added to the data.



```{r}
df %>% 
  select(time, char_time, lat_zero, lon_zero, distance)
```

## Summarizing the Data
Now that we have the distance for each point in time, we can summarize the data. In many cases, we do not need the data at a sub-second or second level. We can summarize the data at a minute level using the `ww_minute_sensorlog` function. This function will take the mean over a period of seconds, usually 60:

```{r summarize_sensorlog_minute}
df_min = ww_minute_sensorlog(df)
head(df_min)
```

We can also summarize the data at a daily level using the `ww_summarize_sensorlog` function:

```{r summarize_sensorlog}
df_sum = ww_summarize_sensorlog(df)
df_sum
```

One of the main issues with daily summaries is that you often want to cross-reference GPS with activity. For that reason, we will keep the data at a minute level for now.


## Reading in the GT3X data
To summarize the activity profile for different levels of proximity to a fixed point, we will read in the GT3X data. This data was captured using an ActiGraph GT9X device in Pensacola, FL. The data is stored in the `extdata` folder of the package, similar to the SensorLog data. We get the path via `ww_example_gt3x_file`:

```{r ag_file}
file_gt3x = ww_example_gt3x_file()
file_gt3x
```


We can read in the data using the `read.gt3x` package, but waterways wraps this using `ww_read_gt3x`.


```{r ag_read}
ag_tz_applied = ww_read_gt3x(file_gt3x, verbose = FALSE, apply_tz = TRUE)
head(ag_tz_applied)
```

We can check the timezone of the data using the `lubridate::tz` function:
```{r ag_tz}
lubridate::tz(ag_tz_applied$time)
```

The `apply_tz` argument in `ww_read_gt3x` will, by default, find the timezone in the header of the file and apply it to the data. This is done using the `lubridate::with_tz` function. The timezone is stored in the file header, which can be seen using the `header` attribute:

We can confirm this via the header:
```{r}
header = attributes(ag_tz_applied)$header
header
header$TimeZone
```

The timezone is stored in the `TimeZone` field. So the data is in GMT, but it is "correct" because this is the local time projected into GMT. For example, if you measured data in a timezone of `-04:00` and `read.gt3x` gave you a value of `2025-03-25 14:14:00`, then the true time in GMT would be `2025-03-25 18:14:00`. Using `read.gt3x`, the time would incorrectly be `2025-03-25 14:14:00 GMT`, but with `ww_read_gt3x` it would be `2025-03-25 18:14:00 GMT`. This means that if you want to convert timezones, you will need to use `lubridate::with_tz` or `lubridate::force_tz`.

We will not apply the timezone information here for the same reasons discussed above for SensorLog. We can read in the data without applying the timezone, giving "incorrect" data, but that can be merged with SensorLog and does not shift times to different dates:

```{r ag_read2}
ag = ww_read_gt3x(file_gt3x, verbose = FALSE, apply_tz = FALSE)
head(ag)
lubridate::tz(ag$time)
```


In ActiLife, the ActiGraph software, values of zero for all axes are set to the last observation carried forward (LOCF). The `fill_zeroes` argument will apply this by default. Also, when the data is read in using `read.gt3x`:
>  DateTimes are therefore represented as POSIXct format with the 'GMT' timezone attribute, which is false; the datetime actually represents local time.


We can calculate a number of summary measures from the acceleration data. We will calculate ActiGraph Activity Counts using the `agcounts` package. We will use `epoch = 60` to calculate the counts in 60-second epochs. From these counts, we can estimate non-wear using the Choi [@choi2011validation] algorithm, which we use from the `actigraph.sleepr` package. The `ww_process_gt3x` function performs these operations together:


```{r calculate_counts}
counts = ww_process_gt3x(ag, verbose = FALSE)
counts %>% 
  head()
counts = counts %>% select(time, counts, wear)
```

### Overlap with SensorLog data
The SensorLog data used here was a subset of the ActiGraph data, and the participant was recorded multiple times. We can see the overlap by using the `time` column.

```{r}
data = counts %>% 
  mutate(in_counts = TRUE) %>% 
  full_join(df_min %>% 
              mutate(in_sensorlog = TRUE),
            by = join_by(time)) %>% 
  tidyr::replace_na(list(in_counts = FALSE, in_sensorlog = FALSE))
data = data %>% 
  filter(in_counts & in_sensorlog)
data
```

Here we see that, for this data, there is complete wear:

```{r}
all(data$wear)
```


Here we can look at the counts within "home"/work and compare them with those outside that area (there are none), as well as a variant that takes wear time into account:
```{r}
data %>% 
  group_by(is_within_home) %>% 
  summarise(
    sum_counts = sum(counts),
    mean_counts = mean(counts),
    sum_counts_wear = sum(counts[wear]),
    mean_counts_wear = mean(counts[wear])
  )
```

### Getting Times
You can easily manipulate the time variables to get date-, hour-, and minute-level data, since you often want to plot the data by hour or minute and facet by day. Here is an example function, `create_date_hour_minute`:

```{r}
create_date_hour_minute = function(data) {
  data = data %>%
    mutate(
      date = lubridate::floor_date(time, "1 day"),
      date = lubridate::as_date(date),
      hour = lubridate::floor_date(time, "1 hour"),
      hour = hms::as_hms(hour),
      minute = lubridate::floor_date(time, "1 minute"),
      minute = hms::as_hms(minute)
    )
}
```


Alternatively, if you group the data by participant, you can create an observation `day` variable that puts data on the same per-person time frame (time from start) versus calendar time:
```{r}
create_day = function(data) {
  data = data %>%
    dplyr::mutate(
      day = as.numeric(difftime(date, min(date), units = "days") + 1)
    )
}
```

The `actibase::acti_separate_times` function performs these steps if the `actibase` package is available.


We can plot the counts over time, colored by whether a point is within home/work (all points are within that region here):
```{r, fig.width = 10, out.width = "100%", fig.asp = 0.618}
if (rlang::is_installed("ggplot2")) {
  library(ggplot2)
  data %>% 
    create_date_hour_minute() %>% 
    ggplot(aes(x = minute, y = counts)) +
    geom_point(aes(colour = is_within_home)) + 
    geom_line() + 
    facet_wrap( ~ date)
}
```

And we can also color by distance from home, so there is not much difference here:
```{r, fig.width = 10, out.width = "100%", fig.asp = 0.618}
if (rlang::is_installed("ggplot2")) {
  library(ggplot2)
  data %>% 
    create_date_hour_minute() %>% 
    ggplot(aes(x = minute, y = counts)) +
    geom_point(aes(colour = distance)) + 
    geom_line() + 
    facet_wrap( ~ date)
}
```


### Estimating Wear Time

We can estimate wear time using the `wear` column from the ActiGraph data. Here we will calculate the sum of wear at each date:

```{r}
data %>% 
  create_date_hour_minute() %>% 
  group_by(date) %>% 
  summarise(
    n = n(),
    n_wear = sum(wear),
    n_nonwear = sum(!wear)
  )
```


Without regard to SensorLog data, we can calculate wear time by date. This is useful if you want to see how much data was collected on a given day:

```{r}
counts %>% 
  create_date_hour_minute() %>% 
  group_by(date) %>% 
  summarise(
    n = n(),
    n_wear = sum(wear),
    n_nonwear = sum(!wear)
  )
```

Once this is calculated, you can exclude dates with insufficient wear time (e.g. less than 95% of the day, or 1368 minutes, or 10 hours of wear). Different tasks require different criteria for wear time.




# References
